Complementing theory on adolescent emotion regulation using  machine learning

### 2022-03-06 manuscript/manuscript.Rmd
minor comments
minor comments after update

abstract: |
  This paper set out to identify the most important early early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests. Candidate predictors were selected based on a prior text mining systematic review of 6305 papers. Participants were 497 Dutch adolescents. Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model. SEM-forests were used to group adolescents with similar trajectories, based on their scores on the predictors. This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories. The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors. Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use. The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.

@@

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to not only complete the cycle, but to enable confirmatory prediction testing growing from exploratory approaches.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

Another relevant insight for theory formation is that,
with the exception of autonomy support and negative parenting,
other parenting behaviors ranked low in predictive importance.
This might seem counter intuitive in light of prior theory and empirical research.
As parents have a legal and moral imperative to care for their children,
much of developmental research has focused on parents' role.
One of the few highly specific theories of emotion regulation development focuses on the family context shapes adolescent emotion regulation development [@morrisRoleFamilyContext2007].
At the same time, however, this finding is consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence [@vanlissaRoleFathersMothers2019; @mastrotheodorosFamilyFunctioningAdolescent2020].
Of course, an alternative explanation might be that parenting effects take place earlier in development,
and are engrained in individual differences by the time youth reach adolescence.
However, a recent study tested this hypothesis in a younger sample and similarly found little evidence for parent-directed processes [@vanlissaMothersFathersQuantitative2020].
<!-- aren’t individual differences the result of the environment where those people were growing up, hence parental style as well? -->
We must thus consider the possibility that parents' influence, at least in adolescence, is less prominent than once was thought.
Future theories of adolescent emotion regulation development should thus at least go beyond parents' role.

@@

The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development, where prior work has taken more piecewise approaches, considering only a few predictors at a time.
This allows us, for the first time, to assess the relative importance of different (themes of) predictors, where prior work has taken a more fragmented approach that does not allow for such comparisons.
Another important contribution of the present study is that, in contrast to many other approaches,
our method accommodates non-linear associations and interactions.
Results indeed showed consistent evidence for non-linear effects,
and some indications for interactions. 
The present study answers recent calls for more person-centered research that explores heterogeneity in developmental processes.
Crucially, whereas other approaches have merely *explored* heterogeneity in outcomes (e.g., using latent class analyses),
the present study *explained* heterogeneity using a comprehensive battery of predictors.
This answers the question why some adolescents experience more difficulties in emotion regulation than others.
Another contribution is that the results have clear implications for theory, assessment, and intervention.
<!-- The present study set out to identify the most important early indicators of developmental trajectories -->
<!-- of emotion regulation across adolescence. -->
<!-- This involves three important challenges: -->
<!-- First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large. -->
<!-- Second, there is a paucity of relevant theory to narrow down the candidate predictors. -->
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
<!-- Machine learning methods that perform variable selection can overcome these challenges. -->
<!-- We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence. -->
<!-- The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development. -->
<!-- The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories. -->

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous  sample from the Netherlands.
This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is made available in a reproducible archive to enable easy replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.

>>>>

abstract: |
  This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests. Candidate predictors were selected based on a prior text mining systematic review of 6305 papers. Participants were 497 Dutch adolescents. Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model. SEM-forests were used to group adolescents with similar trajectories, based on their predictor scores. This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents may follow similar trajectories. The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors. Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use. The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.

@@

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to not only complete the cycle, but to enable tests confirmatory prediction following from exploratory approaches.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

Another relevant insight for theory formation is that,
with the exception of autonomy support and negative parenting,
other parenting behaviors ranked low in predictive importance.
This might seem counter-intuitive in light of prior theory and empirical research.
As parents have a legal and moral imperative to care for their children,
much of developmental research has focused on parents' role.
One of the few highly specific theories of emotion regulation development focuses on the family context shapes adolescent emotion regulation development [@morrisRoleFamilyContext2007].
At the same time, however, this finding is consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence [@vanlissaRoleFathersMothers2019; @mastrotheodorosFamilyFunctioningAdolescent2020].
Of course, an alternative explanation might be that parenting effects take place earlier in development,
and are engrained in individual differences by the time youth reach adolescence.
However, a recent study tested this hypothesis in a younger sample and similarly found little evidence for parent-directed processes [@vanlissaMothersFathersQuantitative2020].
<!-- aren’t individual differences the result of the environment where those people were growing up, hence parental style as well? -->
We must thus consider the possibility that parents' influence, at least in adolescence, is less prominent than once was thought.
Future theories of adolescent emotion regulation development should thus at least go beyond parents' role.

@@

The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development. In contrast, prior work has taken more piecewise approaches, considering only a few predictors at a time.
This allows us, for the first time, to assess the relative importance of different (themes of) predictors, where the more fragmented approach used in prior work does not allow for such comparisons.
Another important contribution of the present study is that, in contrast to many other approaches,
our method accommodates non-linear associations and interactions.
Results indeed showed consistent evidence for non-linear effects,
and some indications for interactions. 
The present study answers recent calls for more person-centered research that explores heterogeneity in developmental processes.
Crucially, whereas other approaches have merely *explored* heterogeneity in outcomes (e.g., using latent class analyses),
the present study *explained* heterogeneity using a comprehensive battery of predictors.
This answers the question why some adolescents experience more difficulties in emotion regulation than others.
Another contribution is that the results have clear implications for theory, assessment, and intervention.
<!-- The present study set out to identify the most important early indicators of developmental trajectories -->
<!-- of emotion regulation across adolescence. -->
<!-- This involves three important challenges: -->
<!-- First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large. -->
<!-- Second, there is a paucity of relevant theory to narrow down the candidate predictors. -->
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
<!-- Machine learning methods that perform variable selection can overcome these challenges. -->
<!-- We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence. -->
<!-- The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development. -->
<!-- The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories. -->

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world. The WEIRD nature of the sample constrict us in generalizing beyond such populations from these findings.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is made available in a reproducible archive to enable easy replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.
